Finally Tagless Staged Interpreters for Simpler Typed Languages

### 2007-04-06 tagless/tagless-final.tex
a few small changes

Our initial motivation came from several papers 
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} which used
embedded interpreters as a central example to justify looking 
into advanced type systems.  While we admire all this technical work,
there remained the issue that we were not convinced that their 
motivating example actually needed all this machinery.

@@

\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
\citet{peyton-jones-simple} seem to argue as follows that a self\hyp
interpreter of a typed language cannot be Jones\hyp optimal.
\begin{enumerate*}
\item One needs to encode a typed language in a typed language based on
a sum type (at some level of the hierarchy),
\item It is not possible to write a \emph{direct} interpreter 
for such an encoding of a typed language
in a typed language, without either an
advanced type system or the overhead of tagging and untagging.
\item Thus an indirect interpreter is necessary, which needs a universal
  type (and hence tagging),
\item Thus any self-interpreter must have tags.
\item Hence the self-interpreter cannot be Jones-optimal.
\end{enumerate*}
While the logic is sound, we showed that the premise in the very first step
was not valid.

\citet{Danvy-tagging-encoding} discuss Jones optimality at length.
There, the authors demonstrate the usefulness of HOAS for typed
self-interpretation.  However, no claim is made that the source program
is itself typed -- just that the interpreter is.  Our reading is that
the source language is not typed.  However, in sharp contrast with 
their interpreter, none of our evaluators can product pattern-match errors
(unlike the code that accompanies \citet{Danvy-tagging-encoding}).
In other words, their language (and interpreter) does use tags, however
the shift to HOAS allows the partial evaluator to remove all the tags.
In contrast, we have no tags at all.

We used a method from \citet{asai-binding-time} to reduce the amount
of back-and-forth between reify/reflect: in our PE we track both
static and dynamic values in parallel (whenever possible).  In some sense,
we are observing that this method type-checks in Hindley-Milner once we
deforest the static code representation.  Put another way, we have
manually applied type-level partial-evaluation to our typelevel 
functions (see \S\ref{S:PE-solution}) to obtain simpler types 
acceptable to MetaOCaml.  It is worthwhile to note that 
\citet{sumii-hybrid} also takes advantage of this method
\citep[of][]{asai-binding-time} to get interesting results combining online
and offline partial evaluation.  Like \citet{SperberThiemann:TwoForOne},
we strive for modularity by reusing earlier stages.  It would be interesting
to see if we could derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
in the same manner.

>>>>

Our initial motivation came from several papers 
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} which used
embedded interpreters as a central example to justify looking 
into advanced type systems.  While we admire all this technical work,
we were not convinced that their 
motivating example actually needed all this machinery.

@@

\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
\citet{peyton-jones-simple} seem to argue as follows that a self\hyp
interpreter of a typed language cannot be Jones\hyp optimal.
\begin{enumerate*}
\item One needs to encode a typed language in a typed language based on
a sum type (at some level of the hierarchy),
\item It is not possible to write a \emph{direct} interpreter 
for such an encoding of a typed language
in a typed language, without either an
advanced type system or the overhead of tagging and untagging.
\item Thus an indirect interpreter is necessary, which needs a universal
  type (and hence tagging),
\item Thus any self-interpreter must have tags and hence cannot be 
  Jones-optimal.
\end{enumerate*}
While the logic is sound, we showed that the premise in the very first step
was not valid.

\citet{Danvy-tagging-encoding} discuss Jones optimality at length and
demonstrate the usefulness of HOAS for typed self-interpretation.
However, no claim is made that the source program is itself typed --
just that the interpreter is.  Our reading is that the source language
is not typed. Therefore, object term encoding has tags and the
interpreter (given in the code accompanying
\citet{Danvy-tagging-encoding}) extensively uses partial
pattern-matching. The shift to HOAS, however, lets the partial
evaluator to remove all the tags. In contrast, our object encoding and
evaluators do not have tags to start with and have no possibility of
pattern-match errors.

We used a method from \citet{asai-binding-time} to reduce the amount
of back-and-forth between reify/reflect: in our PE we track both
static and dynamic values in parallel.  In some sense,
we are observing that this method type-checks in Hindley-Milner once we
deforest the static code representation.  Put another way, we have
manually applied type-level partial-evaluation to our typelevel 
functions (see \S\ref{S:PE-solution}) to obtain simpler types 
acceptable to MetaOCaml.  It is worthwhile to note that 
\citet{sumii-hybrid} also takes advantage of this method
\citep[of][]{asai-binding-time} to get interesting results combining online
and offline partial evaluation.  Like \citet{SperberThiemann:TwoForOne},
we strive for modularity by reusing earlier stages.  It would be interesting
to see if we could derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
in the same manner.
